---
title: "ETC3250/5250: Introduction to Machine Learning"
---

```{r setup, include = FALSE}
# please add any R-packages used here in DESCRIPTION imports
library(yaml)
library(stringr)
library(tibble)
library(dplyr)
library(purrr)
library(glue)
library(knitr)
library(kableExtra)
knitr::opts_chunk$set(echo = FALSE)
```

```{r fns}
make_rectangular <- function(.data) {
  l <- map(.data, ~map(.x, as.character))
  bind_rows(map_dfr(l, ~tibble(!!!.x)))
}


ui_list <- function(vec) {
  for(x in vec) {
    cat("-", x, "\n")
  }
}

```


```{r staffs}
dat <- read_yaml("_course.yml")
team_df <- make_rectangular(dat$team)
schedule_df <- make_rectangular(dat$schedule)

nlecturer <- sum(team_df$role=="lecturer")
ntutor <- sum(team_df$role=="tutor")
```


## Lecturer`r ifelse(nlecturer > 1, 's', '')`

```{r lecturers, results="asis"}
team_df %>% 
  filter(role=="lecturer") %>% 
  pull(name) %>% 
  ui_list()
```

## Tutor`r ifelse(ntutor > 1, 's', '')`

```{r tutors, results="asis"}
team_df %>% 
  filter(role=="tutor") %>% 
  pull(name) %>% 
  ui_list()
```

## Consultations

All consultations will be in using zoom. Check Moodle for the links.

```{r consult, results="asis"}
team_df %>% 
  filter(!is.na(consult)) %>% 
  mutate(name = gsub(" (Chief Examiner)", "", name, fixed = T)) %>% 
  mutate(consult_print = glue::glue("**{name}**: {consult}")) %>% 
  pull(consult_print) %>% 
  ui_list()
```

## Assessments

Final exam 60%, project 15%, each assignment is 10%, weekly quizzes 5%.

## Tentative Schedule

* Lectures: **Mon 2-3pm, Tue 12-1** on zoom (see Moodle for link)
* Tutorials:
  * A: **Mon 17:00** CL_Anc-19.LTB_G21 (Patrick)
  * B: **Mon 19:00** CL_Inn-33.SciPort_GG01 (Patrick)
  * C: **Mon 19:00** on zoom (Shin)
  * D: **Tue 08:00** CL_Chn-20.Menzies_H1 (Kenji)
  * E: **Tue 08:00** on zoom (Patrick)
  * F: **Thu 18:00** CL_Anc-15.Law_G07 (Kenji)
  * G: **Fri 16:00** CL_Anc-19.LTB_393 (Kenji)
  * H: **Fri 18:00** CL_Inn-33.SciPort_G101 (Kenji)
  * I: **Fri 18:00** on zoom (Shin)

There are no lectures, tutorials or consultations during the mid-semester break. You are expected to attend the lecture and ONE tutorial stream. Please see your timetable which tutorial stream you are scheduled in.

```{r}
schedule_df <- schedule_df %>% 
  mutate(weekno = as.integer(str_extract(week, "^[0-9]+"))) %>% 
  mutate(lecture_a = file.exists(sprintf("lectures/lecture-%.2da.html", weekno)),
         lecture_b = file.exists(sprintf("lectures/lecture-%.2db.html", weekno)),
         tutorial = file.exists(sprintf("tutorials/tutorial-%.2d.html", weekno)),
         solution = file.exists(sprintf("tutorials/tutorial-%.2dsol.html", weekno))) %>%  
  mutate(slides = case_when(
    lecture_a & lecture_b ~ glue("A: ","<a class='fa fa-link' href='lectures/{sprintf(\"lecture-%.2dA.html\", weekno)}' target='_blank'></a> <a class='fab fa-r-project' href='lectures/{sprintf(\"lecture-%.2dA.Rmd\", weekno)}' target='_blank'></a> <a class='fa fa-file-pdf' href='lectures/{sprintf(\"lecture-%.2dA.pdf\", weekno)}' target='_blank'></a>"," ; ", "B: ","<a class='fa fa-link' href='lectures/{sprintf(\"lecture-%.2dB.html\", weekno)}' target='_blank'></a> <a class='fab fa-r-project' href='lectures/{sprintf(\"lecture-%.2dB.Rmd\", weekno)}' target='_blank'></a> <a class='fa fa-file-pdf' href='lectures/{sprintf(\"lecture-%.2dB.pdf\", weekno)}' target='_blank'></a>"),
    lecture_a ~ glue("A: ","<a class='fa fa-link' href='lectures/{sprintf(\"lecture-%.2dA.html\", weekno)}' target='_blank'></a> <a class='fab fa-r-project' href='lectures/{sprintf(\"lecture-%.2dA.Rmd\", weekno)}' target='_blank'></a> <a class='fa fa-file-pdf' href='lectures/{sprintf(\"lecture-%.2dA.pdf\", weekno)}' target='_blank'></a>"),
    TRUE ~ ""
  ),
  tut = case_when(
    tutorial & solution ~ glue("<a class='fa fa-tasks' href='tutorials/{sprintf(\"tutorial-%.2d\", weekno)}.html' target='_blank'></a> <a class='fa fa-wrench' href='tutorials/{sprintf(\"tutorial-%.2d\", weekno)}.Rmd' target='_blank'></a> <a class='fa fa-lightbulb' href='tutorials/{sprintf(\"tutorial-%.2d\", weekno)}sol.html' target='_blank'></a> <a class='fa fa-laptop' href='tutorials/{sprintf(\"tutorial-%.2d\", weekno)}sol.Rmd' target='_blank'></a>"),
    tutorial ~ glue("<a class='fa fa-tasks' href='tutorials/{sprintf(\"tutorial-%.2d\", weekno)}.html' target='_blank'></a> <a class='fa fa-wrench' href='tutorials/{sprintf(\"tutorial-%.2d\", weekno)}.Rmd'  target='_blank'></a>"),
    TRUE ~ ""
  ),
  topic = ifelse(is.na(topic), "", topic),
  readings = ifelse(is.na(readings), "", readings),
  assessment = ifelse(is.na(assessment), "", assessment)) %>% 
  select(weekno, slides, tut, topic, readings, assessment)


midsem_break <- 5 + 0 # after which week # needs the fix after week 0 was added

schedule_df %>% 
  kable(escape = FALSE, align = c("c", "c", "c", "l", "l", "l"),
        col.names = c("Week", "Slides", "Tutorial", 
                      "Topic", "Readings", "Assessments")) %>% 
  kable_styling(bootstrap_options = "striped",
                fixed_thead = T) %>% 
  column_spec(1, width = "0.5cm") %>% 
  column_spec(2, width = "1cm") %>% 
  column_spec(3, width = "1cm") %>% 
  column_spec(6, width = "6cm") %>% 
  group_rows("Midsemester Break (1 week)", midsem_break + 1, 12, label_row_css = "background-color: #d9e1e8;") %>% 
  add_indent(1:midsem_break) # this needs to be done to match up with above
```

## Class data sets

- [nrc.csv](data/nrc.csv)
- [chocolates.csv](data/chocolates.csv)

## Expectations

* Unless your question is a personal issue or an administrative matter, please do not send us direct email. The moodle discussion forum is appropriate for questions related to the course.
* For personal or private administrative issues, the email contact is: ETC3250.Clayton-x@monash.edu
* The computer software R (4.0.4 -- 2021-02-15 -- "Lost Library Book") and RStudio Desktop (version 1.4) will be used for the unit. Please install the latest versions on your computer.
* Lectures will be delivered live on zoom, and you are expected to have either attended the lecture, or watched the recordings fully, **prior to the tutorial** for the week.
* Tutorials are not recorded, and attendance is expected.
* The on-campus labs have computers, but we recommend bringing your own laptop with the software installed. Check the tutorial instructions ahead to get the list of software. (If you do not have a laptop you can borrow one from MoVE, and we recommend that you have R/RStudio and the packages necessary on a memory stick to attach to the borrowed laptop.)
* If you register after the start of the semester or if you miss a lecture/tutorial, it is your responsibility to catch up with missed material, learn about due dates for material to be turned in, and getting assigned to a group for team work, as necessary.
* As a Monash unit, instructors and students agree to adhere to [Assessment and Academic Integrity Policy](https://www.monash.edu/__data/assets/pdf_file/0019/2300923/Assessment-and-Academic-Integrity-Policy.pdf). Students are responsible for their own good academic practice and must:
    - undertake their studies and research responsibly and with honesty and integrity;
    - credit the work of others and seek permission to use that work where required;
    - not plagiarise, cheat or falsify their work;
    - ensure that their work is not falsified;
    - not resubmit any assessment they have previously submitted, without the permission of the chief examiner; appropriately acknowledge the work of others;
    - take reasonable steps to ensure that other students are unable to copy or misuse their work; and
    - be aware of and comply with University regulations, policies and procedures relating to academic integrity.

# R package installation list

Here is the list of R packages as of the start of semester. There may be more packages to install as the semester progresses. You should be able to install by cutting and pasting this code into your RStudio console.

```
install.packages(c("knitr", "tidyverse", "tidymodels", "gapminder", "gridExtra", "patchwork", "tourr", "kableExtra", "fpc", "statquotes", "yardstick", "ISLR", "broom", "mvtnorm", "ggdendro", "RColorBrewer", "plotly", "htmltools", "splines", "grid", "rpart", "mgcv", "gratia", "GGally", "ggpubr", "MASS", "caret", "modelr", "boot", "gganimate", "ggrepel", "ggthemes", "datasauRus", "geozoo", "ggparallel", "ggmosaic", "nullabor", "rpart.plot", "e1071", "viridis", "randomForest", "mlbench", "gtable", "keras", "nnet", "penalizedLDA", "ggExtra", "magick"))
```
