---
title: "ETC3250/5250 Introduction to Machine Learning"
---

```{r}
#| label: load_packages
#| include: false
#| message: false
#| warning: false
#| echo: false
#| cache: false
library(tidyverse)
options(knitr.kable.NA = '')
source(here::here("course_info.R"))
week <- as.integer(1 + ((Sys.Date() - as_date(start_semester))/7))
```

## Lecturer/Chief Examiner

* **Professor Di Cook**
    - Email: [etc3250.clayton-x@monash.edu](mailto:etc3250.clayton-x@monash.edu) 
    - Consultation: 

## Tutors

* Patrick Li
    - Tutorials: 
    - Consultation:
* Harriet Mason
    - Tutorials: 
    - Consultation:
* Jayani Lakshika
    - Tutorials: 
    - Consultation:
* Krisanat Anukarnsakulchularp
    - Tutorials: 
    - Consultation:

## Weekly schedule

* Lecture: Wed 1:05-2:45pm
* Tutorial: 1.5 hours
* Weekly learning quizzes due Mondays 9am

```{r}
#| label: schedule2
#| message: false
#| warning: false
#| echo: false
#| output: asis
schedule |>
  transmute(
    Show = !is.na(Week), #& (Week <= week | Week <= 1),
    Topic = if_else(!Show, Topic, glue::glue("[{Topic}](./week{Week}/index.html)")),
    Reference = if_else(is.na(Week), Reference, glue::glue("[{Reference}]({Reference_URL})")),
    Assessments = if_else(is.na(Assignment), Assignment, glue::glue("[{Assignment}]({File})")),
    Week = format(Date, "%d %b")
  ) |>
  select(-Show) |>
  select(Week, everything()) |>
  knitr::kable(format = "markdown")
```

## Assessments

* [Weekly learning quizzes](https://learning.monash.edu/course/view.php?id=9453&section=20): 5%
* [Assignment 1](https://learning.monash.edu/course/view.php?id=9453&section=20): 10%
* [Assignment 2](https://learning.monash.edu/course/view.php?id=9453&section=20): 10%
* [Assignment 3](https://learning.monash.edu/course/view.php?id=9453&section=20): 10%
* [Project](https://learning.monash.edu/course/view.php?id=9453&section=20): 10%
* Final exam: 55%

## R package installation

Here is the code to install (most of) the R packages we will be using in this unit.

```r
install.packages(c("tidyverse", "tidymodels", "tourr", "geozoo", "mulgar", "ggpcp", "plotly", "detourr", "langevitour", "ggbeeswarm", "MASS", "GGally", "ISLR", "rpart", "rpart.plot", "randomForest", "e1071", "Rtsne", "classifly", "penalizedLDA", "nnet", "kernelshap", "shapviz", "iml", DALEX", "cxhull", "fpc", "mclust", "ggdendro", "kohonen", "aweSOM", "patchwork", "ggthemes", "colorspace", "palmerpenguins"), dependencies = TRUE)
```

In addition, follow [these instructions](https://tensorflow.rstudio.com/install/) to set up tensorflow and keras, which requires having python installed. 
