---
title: "ETC3250/5250 Tutorial 9 Instructions"
subtitle: "Model choice, and regularisation"
author: "prepared by Professor Di Cook"
date: "Week 9"
output:
  html_document:
    after_body: tutorial-footer.html
    css: tutorial.css
---

```{r, echo = FALSE, message = FALSE, warning = FALSE, warning = FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE, 
  collapse = TRUE,
  comment = "#",
  fig.height = 4,
  fig.width = 8,
  fig.align = "center",
  cache = FALSE
)
library(emo)
```


```{r, echo = FALSE, message = FALSE, warning = FALSE, warning = FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE, 
  collapse = TRUE,
  comment = "#",
  fig.height = 4,
  fig.width = 8,
  fig.align = "center",
  cache = FALSE
)
```

## `r emo::ji("target")` Objective

The objectives of this tutorial are to

- conduct variable selection and examine importance
- practice methods for choosing models
- fit a regularised model

## `r emo::ji("wrench")` Preparation 

Make sure you have these packages installed:

```
install.packages(c("tidyverse", "ISLR", "leaps", "patchwork", "rsample", "tidymodels", "glmnet", "skimr"))
```

### 1. Choosing variables

Work your way through the textbook lab 6.5.1 Best Subset Selection, and the forward stepwise procedure in 6.5.2, then answer these questions. (Note that there is no tidymodels option for either of these options at the moment.)


a. The `regsubsets()` function (part of the `leaps` library) performs best subset selection by identifying the best model that contains a given number of predictors, where best is quantified using RSS. By default it only examines up to 8 variable models. Which variables make the best 8 variable model?

```{r}
library(knitr)
library(tidyverse)
library(ISLR)
library(skimr)
library(leaps)
library(patchwork)
library(rsample)
library(tidymodels)
library(glmnet)
data(Hitters)
```

```{r eval=FALSE}
# Take a look at the data
skim(Hitters)
```

```{r}
# Handle the missing values on Salary,
# by removing them
Hitters <- Hitters %>%
  filter(!is.na(Salary))

# Subset selection
regfit.full <- regsubsets(Salary~., Hitters)
```

```{r eval=FALSE}
summary(regfit.full)
```

```{r}
coef(regfit.full, 8)
```

b. Set the max number of variables to be 19, by adding the argument `nvmax=19`. Plot the model fit diagnostics for the best model of each size. What would these diagnostics suggest about an appropriate choice of models? Do your results compare with the text book results? Why not?

```{r}
regfit.full <- regsubsets(Salary~., data=Hitters, nvmax=19)
reg.summary <- summary(regfit.full)
```

```{r}
#names(reg.summary)
models <- tibble(nvars=1:19, rsq=reg.summary$rsq, 
                 rss=reg.summary$rss, 
                 adjr2=reg.summary$adjr2, 
                 cp=reg.summary$cp, 
                 bic=reg.summary$bic)
p1 <- ggplot(models, aes(x=nvars, y=rsq)) + geom_line()
p2 <- ggplot(models, aes(x=nvars, y=rss)) + geom_line()
p3 <- ggplot(models, aes(x=nvars, y=adjr2)) + geom_line()
p4 <- ggplot(models, aes(x=nvars, y=cp)) + geom_line()
p5 <- ggplot(models, aes(x=nvars, y=bic)) + geom_line()
p1 + p2 + p3 + p4 + p5
```

c. Fit forward stepwise selection. How would the decision about best model change?  

```{r}
regfit.fwd <- regsubsets(Salary~., data=Hitters, nvmax=19, method="forward")
#summary(regfit.fwd)

d <- tibble(nvars=0:19,  
                 rss=regfit.fwd$rss)
ggplot(d, aes(x=nvars, y=rss)) + geom_line()
```

### 2. Training and testing sets with variable selection

a. Break the data into a 2/3 training and 1/3 test set. 
b. Fit the best subsets. Compute the mean square error for the test set. Which model would it suggest? Is the subset of models similar to produced on the full data set? Do your results compare with the text book results? Why not?


```{r}
set.seed(1)
split <- initial_split(Hitters, 2/3)
h_tr <- training(split)
h_ts <- testing(split)

regfit.best <- regsubsets(Salary~., data=h_tr, nvmax=19)
test.mat <- model.matrix(Salary~., data=h_ts)
val.errors <- rep(NA,19)
for (i in 1:19) {
   coefi <- coef(regfit.best, id=i)
   pred <- test.mat[,names(coefi)]%*%coefi
   val.errors[i] <- mean((h_ts$Salary-pred)^2)
}
val.errors
d2 <- tibble(nvars=1:19,  
                 err = val.errors)
ggplot(d2, aes(x=nvars, y=err)) + geom_line()
coef(regfit.best, which.min(val.errors))
coef(regfit.full, 10)
```

### 3. Cross-validation with variable selection

It is said that 10-fold cross-validation is a reasonable choice for dividing the data. What size data sets would this create for this data? Argue whether this is good or bad. 

```{r eval=FALSE, echo=FALSE}
# This is the code to do cross-validation, if you are interested
# Make our own prediction function
predict.regsubsets <- function(object,newdata,id,...){
  form=as.formula(object$call[[2]])
  mat=model.matrix(form,newdata)
  coefi=coef(object,id=id)
  xvars=names(coefi)
  mat[,xvars]%*%coefi
}
# Create the folds
set.seed(20190513)
k <- 5
folds <- createFolds(Hitters$Salary, k=k)

# Compute errors for different folds
cv.errors <- matrix(NA,k,19, dimnames=list(NULL, paste(1:19)))
for (j in 1:k) {
  best.fit <- regsubsets(Salary~., data=Hitters[(1:263)[folds[[j]]],], nvmax=19)
  for(i in 1:19) {
    pred <- predict(best.fit, Hitters[folds[[j]],], id=i)
    cv.errors[j,i] <- mean((Hitters$Salary[folds[[j]]]-pred)^2)
    }
}
# Put data in long form to plot effectively
cv.errors <- as_tibble(cv.errors) %>%
  gather(nvars, error) %>% mutate(nvars=as.numeric(nvars)) %>%
  mutate(cv_set = rep(1:5, 19))
ggplot(cv.errors, aes(x=nvars, y=error, colour=factor(cv_set))) +
  geom_line() + theme(legend.position="none")
```


### 4. Regularisation for variable selection

Here we will use lasso to fit a regularised regression model and compare the results with the best subset model.

a. Using your results from questions 1-3, fit the best least squares model, to your training set. Write down the mean square error and estimates for the final model. We'll use these to compare with the lasso fit.

```{r}
min(val.errors)
coef(regfit.best, which.min(val.errors))
```

b. Fit the lasso to a range of $\lambda$ values. Plot the standardised coefficients against $\lambda$. What does this suggest about the predictors?


```{r}
grid <- 10^seq(10,-2,length=100)
x <- model.matrix(Salary~., h_tr)[,-1]
y <- h_tr$Salary

lasso.mod <- glmnet(x, y, alpha=1, lambda=grid)
# Need a coefficient matrix of 100(nlambda)x19(p)
# as.matrix function converts sparse format into this
coeffs <- as.matrix(lasso.mod$beta) 
coeffs <- cbind(var=rownames(coeffs), coeffs)
cv <- coeffs %>% as_tibble() %>%
  gather(nval, coeffs, -var) %>%
  mutate(coeffs=as.numeric(coeffs)) %>%
  mutate(lambda=rep(lasso.mod$lambda, rep(19, 100)))
  
p <- ggplot(cv, aes(x=lambda, y=coeffs, group=var, label=var)) + geom_line() +
  scale_x_log10(limits=c(-1, 100))
plotly::ggplotly(p)
# This is how the sample code plots
#plot(lasso.mod, xvar="lambda", xlim=c(-1, 5))
```


c. Now use cross-validation to choose the best $\lambda$.

```{r}
# Do cross-validation, using glmnet's function
set.seed(1)
cv.out <- cv.glmnet(x, y, alpha=1)
#cv.df <- tibble(lambda=cv.out$lambda, mse=cv.out$cvm,
#                mse_l=cv.out$cvlo, mse_h=cv.out$cvup)
#ggplot(cv.df, aes(x=lambda, y=mse)) + geom_point() +
#  scale_x_log10() +
#  geom_errorbar(aes(ymin=mse_l, ymax=mse_h))
# This is how the sample code plots
plot(cv.out)
```


d. Fit the final model using the best $\lambda$. What are the estimated coefficients? What predictors contribute to the model?

```{r}
# Fit a single model to the best lambda, predict the test set, and compute mse
bestlam <- cv.out$lambda.min
bestlam
x_ts <- model.matrix(Salary~., h_ts)[,-1]
y_ts <- h_ts$Salary
lasso.pred <- predict(lasso.mod, s=bestlam, newx=x_ts)
y.test <- y_ts
mean((lasso.pred-y.test)^2)

# Fit the model
# alpha=1 means lasso
# Its strange that it still needs the grid of lambdas to fit
# but it seems necessary for the optimisation.
# Then the predictions for best lambda made with predict fn
out <- glmnet(x, y, alpha=1, lambda=grid)
lasso.coef <- predict(out, type="coefficients", s=bestlam)[1:20,]
lasso.coef
lasso.coef[lasso.coef!=0]
```


e. Does the best lasso model beat the best least squares model (best subsets)? 


### 5. Making sense of it all

Only one variable is very important for the model, which is it? (It occurs in every list of variables returned as important, and has the highest coefficient in the lasso model.) Several more variables frequently appear as important, which ones are these? Several others, appear occasionally in some models but always have very small coefficients. Can you name one of these? What does this tell you about the relationship between Salary and all the predictors?

##### © Copyright 2022 Monash University
