---
title: "Week 1: Foundations of machine learning"
---

```{r}
week <- 1
source(here::here("course_info.R"))
```

## Main reference

```{r}
#| output: asis
schedule |>
  filter(Week == week) |>
  mutate(ref = glue::glue("[{Reference}]({Reference_URL})")) |>
  pull(ref)
```

## What you will learn this week

* Framing the problems
* Notation and math
* Bias variance-tradeoff
* Fitting your models: training/test splits, optimisation
* Measuring fit: accuracy, loss
* Diagnostics: residuals
* Feature engineering: combining variables to better match purpose and help the model fitting

## Lecture slides

- [html](https://iml.numbat.space/week1/slides.html)
- [pdf](https://iml.numbat.space/week1/slides.pdf)
- [quarto](https://iml.numbat.space/week1/slides.qmd)
- [R code](https://iml.numbat.space/week1/slides.R)

## Tutorial instructions

Instructions:

- [html](https://iml.numbat.space/week1/tutorial.html)
- [quarto](https://iml.numbat.space/week1/tutorial.qmd)

## Assignments
```{r}
#| output: asis
show_assignments(week)
```
