---
title: "ETC3250/5250 Tutorial 7 Instructions"
subtitle: "Tree models"
author: "prepared by Professor Di Cook"
date: "Week 7"
output:
  html_document:
    after_body: tutorial-footer.html
    css: tutorial.css
---

```{r, echo = FALSE, message = FALSE, warning = FALSE, warning = FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE, 
  eval = FALSE,
  collapse = TRUE,
  comment = "#",
  fig.height = 4,
  fig.width = 8,
  fig.align = "center",
  cache = FALSE
)
library(emo)
```


## `r emo::ji("target")` Objective

The objectives for this week are to 

- Practice fitting a classification tree model
- Understand the way the tree model is fitted based on impurity
- Learn about the relationship between fitting parameters, bias and variance

## `r emo::ji("wrench")` Preparation 

Make sure you have these packages installed:

```
install.packages(c("tidyverse", "tidymodels", "tourr", "rpart.plot", "discrim"))
```

### `r emo::ji("book")` Reading 

- Textbook sections 8.1, 8.3

## `r emo::ji("waving_hand")` Getting started

If you are in a zoom tutorial, say hello in the chat. If in person, do say hello to your tutor and to your neighbours. 


## `r emo::ji("gear")` Exercises 

### 1. This question is about entropy as an impurity metric  for a classification  tree.

a. Write down the formula for entropy as an impurity measure for two groups. 
b. Establish that the the worst case split has 50% one group and 50% the other group, in whatever way you  would like (algebraicly or graphically).
c. Extend the entropy formula so that it can be used to describe the  impurity for a possible  split of the data into two subsets. That is, it needs  to be the sum of the impurity for both left and right subsets of data.

### 2. Computing impurity

For this sample of data, 
```{r}
df <- tibble(x=c(1,3,4,5,7), y=c("A", "B", "A", "B", "B"))
kable(df) %>% kable_styling()
```
a. Compute the entropy impurity metric for all possible splits.
b. Write  down the classification rule for the tree that would  be formed for the best split.
    
### 4. Write decision tree model

For the following data set, compute the default classification tree. Write out the tree rules, and also sketch the boundary between classes.

a. olive oils,  for three regions

```{r}
olive <- read_csv("http://www.ggobi.org/book/data/olive.csv") %>%
  rename(name=X1) %>%
  dplyr::select(-name, -area) %>%
  mutate(region = factor(region))
```

b. chocolates, for type

```{r}
choc <-  read_csv("data/chocolates.csv") %>%
  select(Type:Protein_g)
```

c. flea, for species

```{r}
data(flea)
```

### 5. Which model should perform best

For the crabs data, make a new variable combining species and gender into one class variable.

a. Use the  grand and guided tour with the LDA index to  examine  the data. Describe the shape. Between LDA and a classification  tree which do  you expect to perform better on this  data?
    
```{r}
crabs <- read_csv("http://www.ggobi.org/book/data/australian-crabs.csv") %>%
  mutate(class = interaction(species, sex)) %>%
  dplyr::select(-index, -species,-sex)
```
b. Use 10-fold cross-validation to determine the best choice of minsplit, for the training set of an 80:20 training:test split of the original data. (Check the code from the lecture 6a/b notes to use as an example.)
c. Fit the classification tree with the recommended minsplit. Compute the test accuracy, using your 20% test set. Explain why the tree is so complicated. Compare with the accuracy from an LDA. Is this consistent with what you thought would be the best model?


##### © Copyright 2021 Monash University
