---
title: "ETC3250/5250 Tutorial 6 Solution"
subtitle: "Visualising high-dimensional data"
author: "prepared by Professor Di Cook"
date: "Week 6"
output:
  html_document:
    after_body: tutorial-footer.html
    css: tutorial.css
---

```{r, echo = FALSE, message = FALSE, warning = FALSE, warning = FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE, 
  collapse = TRUE,
  comment = "#",
  fig.height = 4,
  fig.width = 8,
  fig.align = "center",
  cache = FALSE
)
library(emo)
```


```{r, echo = FALSE, message = FALSE, warning = FALSE, warning = FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE, 
  collapse = TRUE,
  comment = "#",
  fig.height = 4,
  fig.width = 8,
  fig.align = "center",
  cache = FALSE
)
```


## 💬 Class discussion exercises

- In tour 1, how many clusters do you see? [tour1](https://iml.numbat.space/tutorials/tour1.html)

```{r eval=FALSE, echo=TRUE}
# This is code that YOU CAN RUN YOURSELF to see the tour, 
# but its not necessary to run in order to do the exercise 
library(tidyverse)
olive <- read_csv("http://www.ggobi.org/book/data/olive.csv") %>%
  dplyr::select(-X1)
library(tourr)
animate_xy(olive[, 3:10], axes="off")
```

```{r eval=FALSE, echo=TRUE}
# This is code was used to create the animation,
# YOU DO NOT NEED to run in order to do the exercise 
# 
library(plotly)
library(htmltools)
set.seed(20190331)
bases <- save_history(olive[,3:10], grand_tour(2), 
    start=matrix(c(1,0,0,1,0,0,0,0,0,0,0,0,0,0,0,0), ncol=2, byrow=TRUE), 
    max = 15)
# Re-set start bc seems to go awry
bases[,,1] <- matrix(c(1,0,0,1,0,0,0,0,0,0,0,0,0,0,0,0), ncol=2, byrow=TRUE)
tour_path <- interpolate(bases, 0.1)
d <- dim(tour_path)
olive_std <- tourr::rescale(olive[,3:10])
mydat <- NULL; 
for (i in 1:d[3]) {
  fp <- as.matrix(olive_std) %*% matrix(tour_path[,,i], ncol=2)
  fp <- tourr::center(fp)
  colnames(fp) <- c("d1", "d2")
  mydat <- rbind(mydat, cbind(fp, rep(i+10, nrow(fp))))
}
colnames(mydat)[3] <- "indx"
df <- as_tibble(mydat) 
p <- ggplot() +
       geom_point(data = df, aes(x = d1, y = d2, 
                                 frame = indx), size=1) +
       theme_void() +
       coord_fixed() +
  theme(legend.position="none")
pg <- ggplotly(p, width=400, height=400) %>%
  animation_opts(200, redraw = FALSE, 
                 easing = "linear", transition=0)
save_html(pg, file="tour1.html")
```

- In tour 2, where three classes have been coloured, how many additional clusters do you see? [tour2](https://iml.numbat.space/labs/tour2.html)

```{r eval=FALSE, echo=TRUE}
# This is code that YOU CAN RUN YOURSELF to see the tour, 
# but its not necessary to run in order to do the exercise 
animate_xy(olive[,3:10], axes="off", col=olive$region)
```

```{r eval=FALSE, echo=TRUE}
# This is code was used to create the animation,
# YOU DO NOT NEED to run it in order to do the exercise 
df <- df %>%
  mutate(region=factor(rep(olive$region, d[3])))
p <- ggplot() +
       geom_point(data = df, aes(x = d1, y = d2, colour=region,
                                 frame = indx), size=1) +
       scale_colour_brewer("", palette="Dark2") +
       theme_void() +
       coord_fixed() +
  theme(legend.position="none")
pg <- ggplotly(p, width=400, height=400) %>%
  animation_opts(200, redraw = FALSE, 
                 easing = "linear", transition=0)
save_html(pg, file="tour2.html")
```

## `r emo::ji("gear")` Exercises 

### 1. Data with different variance-covariances

Take a look at the data from tutorial 5 using a grand tour. Can you see the difference between the two sets better now? You should see that one group has roughly the same spread of observations for each group, in any combination of the variables. The other has some combinations of the variables where the spread is different for each group. 

**You should be able to see that set B has two groups where the variance-covariance is elliptical in shape but oriented very differently.**

You can use this code to run the tour:

```{r eval=FALSE}
library(tourr)
animate_xy(setA[,1:3], col=setA$class)
animate_xy(setB[,1:3], col=setB$class)
```

**Set A** has equal variance-covariance between groups, $\Sigma$: 

$$\Sigma = \begin{bmatrix} 3.0&0.2&-1.2&0.9\\
0.2&2.5&-1.4&0.3\\
-1.2&-1.4&2.0&1.0\\
0.9&0.3&1.0&3.0\\
\end{bmatrix}$$
 
and **set B** has different variance-covariances between groups, $\Sigma_1, \Sigma_2, \Sigma_3$:

$\Sigma_1 = \Sigma$

$$\Sigma_2 = \begin{bmatrix}3.0&-0.8&1.2&0.3\\
-0.8&2.5&1.4&0.3\\
1.2&1.4&2.0&1.0\\
0.3&0.3&1.0&3.0\\
\end{bmatrix}$$

$$\Sigma_3 = \begin{bmatrix}2.0&-1.0&1.2&0.3\\
-1.0&2.5&1.4&0.3\\
1.2&1.4&4.0&-1.2\\
0.3&0.3&-1.2&3.0\\
\end{bmatrix}$$

This code is used simulate the data:

```{r echo=TRUE, eval=FALSE}
set.seed(20200416)
library(mvtnorm)
vc1 <- matrix(c(3, 0.2, -1.2, 0.9, 0.2, 2.5, -1.4, 0.3, -1.2, -1.4, 2.0, 1.0, 0.9, 0.3, 1.0, 3.0), ncol=4, byrow=TRUE)
vc2 <- matrix(c(3, -0.8, 1.2, 0.3, -0.8, 2.5, 1.4, 0.3, 1.2, 1.4, 2.0, 1.0, 0.3, 0.3, 1.0, 3.0), ncol=4, byrow=TRUE)
vc3 <- matrix(c(2.0, -1.0, 1.2, 0.3, -1.0, 2.5, 1.4, 0.3, 1.2, 1.4, 4.0, -1.2, 0.3, 0.3, -1.2, 3.0), ncol=4, byrow=TRUE)
m1 <- c(0,0,3,0)
m2 <- c(0,3,-3,0)
m3 <- c(-3,0,3,3)
n1 <- 85
n2 <- 104
n3 <- 48
setA <- rbind(rmvnorm(n1, m1, vc1), rmvnorm(n2, m2, vc1), rmvnorm(n3, m3, vc1))
setA <- data.frame(setA)
setA$class <- c(rep("1", n1), rep("2", n2), rep("3", n3))
setB <- rbind(rmvnorm(n1, m1, vc1), rmvnorm(n2, m2, vc2), rmvnorm(n3, m3, vc3))
setB <- data.frame(setB)
setB$class <- c(rep("1", n1), rep("2", n2), rep("3", n3))
```

### 2. Exploring for class separations, heterogeneous variance-covariance and outliers

Remember the chocolates data? The chocolates data was compiled by students in a previous class of Prof Cook, by collecting nutrition information on the chocolates as listed on their internet sites. All numbers were normalised to be equivalent to a 100g serving. Units of measurement are listed in the variable name. You are interested in answering "How do milk and dark chocolates differ on nutritional values?"

a. Examine all of the nutritional variables, relative to the chocolate type, using a grand tour (`tourr::animate_xy()`) and a guided tour (look up the help for `tourr::guided_tour` to see example of how to use the `lda_pp` index). Explain what you see in terms of differences between groups.

**You can see some differences between the two groups, that they are almost separated, and some strong linear associations, in some combinations of variables.** 

**The guided tour with LDA index stops at a projection where most milk chocolates are separated from most dark chocolates, although there is no gap between the two clusters.**

b. From the tour, should you assume that the variance-covariance matrices of the two types of chocolates is the same? Regardless of your answer, conduct a linear discriminant analysis, on the standardised chocolates data. Because the variables are standardised the magnitude of the coefficients of the linear discriminant can be used to determine the most important variables. What are the three most important variables? What are the four least important variables? Look at the data with a grand tour of only the three important variables, and discuss the differences between the groups. 

**The variance-covariance matrices are clearly different between the two groups.**

**The three most important variables are Fiber_g, Na_mg and TotFat_g**

**The least important variables are Calories, CalFat, SatFat_g, Carbs_g.**

**Using just the three most important variables in the tour, you can see the differences between the two groups more clearly.**

c. Filter the data to only focus on the milk chocolates. Explain what you see in terms of association between variables, and outliers. (Note that sometimes the guided tour using the `cmass` index can be useful for detecting multivariate outliers, but it doesn't help here. The grand tour alone is the best for spotting them.)

**Three outliers, two close together, can be seen.**

```{r echo=TRUE, eval=FALSE}
library(tidyverse)
library(tourr)
choc <- read_csv("http://iml.numbat.space/data/chocolates.csv")
std <- function(x) (x-mean(x))/sd(x)
choc_std <- choc %>%
  mutate_if(is.numeric, std)
animate_xy(choc_std[,5:14], col=choc$Type)
animate_xy(choc_std[,5:14], 
           guided_tour(lda_pp(choc_std$Type)), 
           col = choc_std$Type)

# Conduct LDA
library(discrim)
library(parsnip)
library(MASS)

choc_std <- choc_std %>%
  mutate(Type = as.factor(Type))
lda_mod <- discrim_linear() %>% 
  set_engine("MASS", prior = c(0.5, 0.5)) %>% 
  translate()

choc_lda_fit <- 
  lda_mod %>% 
  fit(Type ~ ., 
      data = choc_std[, 4:14])
choc_lda_fit
animate_xy(choc_std[,c(7,10,12)], col=choc$Type)

# Milk chocolates only
animate_xy(choc_std[choc_std$Type == "Milk", 5:14])
animate_xy(choc_std[choc_std$Type == "Milk", 5:14], 
           guided_tour(cmass()))
```


### 3. Assessing variable importance with the manual tour

This example uses the olive oils data. 

a. Read in the data. Keep `region` and the fatty acid content variables. Standardize the variables to have mean and variance 1.
b. Fit a linear discriminant analysis model, to a training set of data. This will produce a 2D discriminant space because there are three variables. Based on the coefficients which variable(s) are important for the first direction, and which are important for the second direction?

**eicosenoic is very important for the first direction, and oleic and palmitic are most important for the second direction.**

c. Using a manual tour, with the `play_manual_tour` function from the `spinifex` package, , starting from the projection given by the discriminant space explore the importance of (i) `eicosenoic` for separating region 1, (ii) `oleic` and `linoleic` for separating regions 2 and 3, (iii) and that `stearic` is not important for any of the separations.

**(i) When eicosenoic is removed from the projection, region 1 observations get mixed/confused with the other regions.**

**(ii) When oleic is removed regions 2 and 3 get confused, BUT when oleic is 100% in the projection, the separation is also a little worse. This means that a combination of oleic and palmitic is better for separating the two groups. It is even more interesting when linoleic contributes more to the projection because the separation between groups is even better. LDA doesn't use gaps between clusters to build the separation, it uses the groups means, and makes the assumption of equal variance-covariance, so it misses this improved solution.**

**(iii) The separation between all groups is better when stearic does NOT contribute to the projection.**

```{r echo=TRUE, eval=FALSE}
library(tidyverse)
library(tourr)
library(rsample)
library(parsnip)
library(discrim)
library(yardstick)
library(spinifex)

olive <- read_csv("http://ggobi.org/book/data/olive.csv") %>%
  # dplyr::filter(region != 1) %>%
  # dplyr::select(region, oleic, linoleic) %>%
  dplyr::select(-X1, -area) %>%
  mutate(region = factor(region))

# Standardise variables
std <- function(x) (x-mean(x))/sd(x)
olive_std <- olive %>%
  mutate_if(is.numeric, std)

set.seed(775)
olive_split <- initial_split(olive_std, 2/3, strata = region)
olive_train <- analysis(olive_split)
olive_test <- assessment(olive_split)

lda_mod <- discrim_linear() %>% 
  set_engine("MASS", prior = c(1/3, 1/3, 1/3)) %>% 
  translate()

olive_lda_fit <- 
  lda_mod %>% 
  fit(region ~ ., 
      data = olive_train)
olive_lda_fit

# Collect the projection giving best separation, 
# to use with manual tour for assessing importance
bas <- olive_lda_fit$fit$scaling

play_manual_tour(bas, olive_std[,2:9], 8, 
                 aes_args = list(color = olive_std$region)) # eicosenoic
play_manual_tour(bas, olive_std[,2:9], 4, 
                 aes_args = list(color = olive_std$region)) # oleic
play_manual_tour(bas, olive_std[,2:9], 5, 
                 aes_args = list(color = olive_std$region)) # linoleic
play_manual_tour(bas, olive_std[,2:9], 3, 
                 aes_args = list(color = olive_std$region)) # stearic
```

##### © Copyright 2021 Monash University
