---
title: "ETC3250/5250 Tutorial 2 Instructions"
subtitle: "Machine learning concepts"
author: "prepared by Professor Di Cook"
date: "Week 2"
output:
  html_document:
    after_body: tutorial-footer.html
    css: tutorial.css
---

```{r, echo = FALSE, message = FALSE, warning = FALSE, warning = FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE, 
  collapse = TRUE,
  comment = "#",
  fig.height = 4,
  fig.width = 8,
  fig.align = "center",
  cache = FALSE
)
library(emo)
```


```{r, echo = FALSE, message = FALSE, warning = FALSE, warning = FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE, 
  collapse = TRUE,
  comment = "#",
  fig.height = 4,
  fig.width = 8,
  fig.align = "center",
  cache = FALSE
)
```

## `r emo::ji("target")` Objective

The objectives for this week are to discuss key concepts for machine learning, practice making summary statistics and plots for a data set, that would be analysed using supervised classification.

## `r emo::ji("wrench")` Preparation 

Make sure you have these packages installed:

```
install.packages(c("tidyverse", "kableExtra", "broom"))
```

### `r emo::ji("book")` Reading

- Read through chapter 2 of the textbook.
- Take a look at the textbook questions: 2.1, 2.2, 2.4. Come up with an answer for each, to discuss with your class mates.
- Work through textbook question 2.7. Bring your answer to class for discussion.


## `r emo::ji("waving_hand")` Getting started

If you are in a zoom tutorial, say hello in the chat. If in person, do say hello to your tutor and to your neighbours. 

## 💬 Class discussion exercises

1. Discuss your answers to 2.1, 2.2, 2.4.
2. What did you get for 2.7?

## `r emo::ji("gear")` Exercise 

Complete these exercises by writing your responses into an Rmarkdown document. 
Give your Rmd file to another group member, and see if they can `knit` it. 

(a) Download the chocolates data set, and read into R (recommend using `read_csv` from the `tidyverse` suite). 

*About the data:* The chocolates data was compiled by students in a previous class of Prof Cook, by collecting nutrition information on the chocolates as listed on their internet sites. All numbers were normalised to be equivalent to a 100g serving. Units of measurement are listed in the variable name.

(b) Take a look at the type of variables in the data. If your question is "How do milk and dark chocolates differ?" what type of problem have you got?
(c) Compute the means and standard deviations for milk and dark on each of the variables. Make a nice table summary. (Try using the pipe operator, with the wrangling verbs `group_by` and `summarise`, and make the table with the `kableExtra` package.)
(d) Make side-by-side boxplots for each of the variables, for type of chocolate. (Use the grammar of graphics in `ggplot2`.) Write a paragraph explaining how the type of chocolate differs nutritionally.
(e) Compute two sample t-tests for each of the variables. Which variable most distinguishes the chocolate type? (This may need to be done using the base R function.)
(f) What sort of difference is the $t$-test is communicating? Is this the same as what we would like to do for building a classifier?

## `r emo::ji("stop_sign")` Wrapping up

Talk to your tutor about what you think you learned today, what was easy, what was fun, what you found hard.
