---
title: "ETC3250/5250 Tutorial 2"
subtitle: "Basics of machine learning"
author: "Prof. Di Cook"
date: "2024-03-04"
quarto-required: ">=1.3.0"
format:
    unilur-html:
        output-file: tutorial.html
        embed-resources: true
        css: "../assets/tutorial.css"
    unilur-html+solution:
        output-file: tutorialsol.html
        embed-resources: true
        css: "../assets/tutorial.css"
unilur-solution: true
---

<!--
devtools::install_github("koncina/unilur")
devtools::install_github("hadley/emo")

Had to run this in the week 1 folder in the terminal in R
cd C:\Users\jjew0003\Documents\Monash Teaching\ETC3250-5250\2025\week2
quarto install extension ginolhac/unilur

or 
cd C:\Users\jjew0003\Documents\Monash_Yr1\ETC3250\S12026\week2
quarto install extension ginolhac/unilur

-->

```{r echo=FALSE}
# Set up chunk for all slides
knitr::opts_chunk$set(
  fig.width = 4,
  fig.height = 4,
  fig.align = "center",
  out.width = "60%",
  code.line.numbers = FALSE,
  fig.retina = 3,
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  dev.args = list(pointsize = 11)
)
```

```{r}
#| echo: false
# Load libraries used everywhere
library(tidyverse)
library(tidymodels)
library(conflicted)
library(colorspace)
library(patchwork)
library(MASS)
library(randomForest)
library(gridExtra)
library(GGally)
library(geozoo)
library(mulgar)
theme_set(theme_bw(base_size = 14) +
   theme(
     aspect.ratio = 1,
     plot.background = element_rect(fill = 'transparent', colour = NA),
     plot.title.position = "plot",
     plot.title = element_text(size = 24),
     panel.background = element_rect(fill = 'transparent', colour = NA),
     legend.background = element_rect(fill = 'transparent', colour = NA),
     legend.key = element_rect(fill = 'transparent', colour = NA)
   )
)
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::select)
conflicts_prefer(dplyr::slice)
conflicts_prefer(palmerpenguins::penguins)
conflicts_prefer(tourr::flea)
```

## `r emo::ji("target")` Objectives

The goal for this week is for you to learn and practice some of the basics of machine learning. 

## `r emo::ji("wrench")` Preparation 

- Complete the quiz
- Do the reading related to week 1


## Exercises: 

Open your project for this unit called `iml.Rproj`. 

#### 1. Answer the following questions for this data matrix, 

$$
\begin{align*}
{\mathbf X} = \left[\begin{array}{rrrrr} 
2 & -2 & -8 & 6 & -7 \\
6 & 6 & -4 & 9 & 6 \\
5 & 4 & 3 & -7 & 8 \\
1 & -7 & 6 & 7 & -1
\end{array}\right]
\end{align*}
$$

a. What is $X_1$ (variable 1)?

::: unilur-solution
$X_1 = (2 ~6 ~5 ~1)$
:::

b. What is observation 3?

::: unilur-solution
$5 ~ 4 ~ 3 ~ -7 ~ 8$
:::

c. What is $n$?

::: unilur-solution
$4$
:::

d. What is $p$?

::: unilur-solution
$5$
::: 

e. What is $X^\top$?

::: unilur-solution
$$
\begin{align*}
{\mathbf X}^\top = \left[\begin{array}{rrrr} 
2 & 6 & 5 & 1\\
-2 & 6 & 4 & -7\\ 
-8 & -4 & 3 & 6 \\
6 & 9 & -7 & 7 \\
-7 & 6 & 8 & -1
\end{array}\right]
\end{align*}
$$
:::

#### 2. Which of these statements is the most accurate? And which is the most precise?

*A. It is almost certain to rain in the next week.*

*B. It is 90% likely to get at least 10mm of rain tomorrow.*

::: unilur-solution
A is more accurate, but B is more precise. 
:::

#### 3. For the following data, make an appropriate training test split of 60:40. The response variable is `cause`. Deomstrate that you have made an appropriate split. 

```{r}
library(readr)
library(dplyr)
library(rsample)

bushfires <- read_csv("https://raw.githubusercontent.com/dicook/mulgar_book/pdf/data/bushfires_2019-2020.csv")
bushfires |> count(cause)
```


::: unilur-solution

The data is unbalanced, so it is especially important to stratify the sampling by the response variable. Without stratifying the test set is likely missing observations in the `burning_off` category. 

```{r}
set.seed(1156)
bushfires_split <- initial_split(bushfires, prop = 0.60, strata=cause)
training(bushfires_split) |> count(cause)
testing(bushfires_split) |> count(cause)
```
:::

#### 4. Consider the following supervised classification data. Answer the questions below for this data.

```{r }
# This code provides a data set to use for the question
library(palmerpenguins)
library(MASS)
p_sub_std <- penguins |>
  dplyr::filter(species != "Gentoo") |>
  rename(
    bl = bill_length_mm,
    bm = body_mass_g,
    fl = flipper_length_mm,
    bd = bill_depth_mm
  ) |>
  dplyr::select(species, bl, bd, fl, bm) |>
  mutate(species = factor(species)) |>
  tidyr::drop_na() |>
  mutate(across(!species, ~ as.numeric(scale(.x))))


```

a. How many classes?

::: unilur-solution
```{r}
p_sub_std |> count(species)
```
:::

b. Create a 80:20 train test split stratifying by species

::: unilur-solution
```{r}
set.seed(1156)
p_sub_std_split <- initial_split(p_sub_std, prop = 0.80, strata=species)
p_sub_std_train <- training(p_sub_std_split)
p_sub_std_test <- testing(p_sub_std_split)
```
:::

c. Fit a kNN model with K = 5 to the training set and predict the test set. Add your first model's probabilistic predictions to the test data

::: unilur-solution

First we specify the kNN model with $K = 5$

```{r}

K1 <- 5

knn_spec_1 <- nearest_neighbor(neighbors = K1, weight_func = "rectangular", dist_power = 2) |> #specify the model type
  set_mode("classification") |>
  set_engine("kknn") #specify how to learn the parameters
```

Then we fit it to data

```{r}

knn_fit_1 <- knn_spec_1 |> ## apply the data
  fit(species ~ ., data = p_sub_std_train)

```

Then we predict the test set

```{r}

p_sub_std_test <- p_sub_std_test |>
  mutate(knn1 = predict(knn_fit_1$fit,
                             p_sub_std_test, type = "prob")[,1])


p_sub_std_test$knn1
```


:::

d. Fit a kNN model with K = 10 to the training set and predict the test set. Add your second model's probabilistic predictions to the test data

::: unilur-solution

First we specify the kNN model with $K = 10$

```{r}

K2 <- 10

knn_spec_2 <- nearest_neighbor(neighbors = K2, weight_func = "rectangular", dist_power = 2) |> #specify the model type
  set_mode("classification") |>
  set_engine("kknn") #specify how to learn the parameters
```

Then we fit it to data

```{r}

knn_fit_2 <- knn_spec_2 |> ## apply the data
  fit(species ~ ., data = p_sub_std_train)

```

Then we predict the test set

```{r}

p_sub_std_test <- p_sub_std_test |>
  mutate(knn2 = predict(knn_fit_2$fit,
                             p_sub_std_test, type = "prob")[,1])


p_sub_std_test$knn2
```


:::

e. Lastly, fit a kNN model with K = 20 to the training set and predict the test set. Add your third model's probabilistic predictions to the test data

::: unilur-solution

First we specify the kNN model with $K = 20$

```{r}

K3 <- 20

knn_spec_3 <- nearest_neighbor(neighbors = K3, weight_func = "rectangular", dist_power = 2) |> #specify the model type
  set_mode("classification") |>
  set_engine("kknn") #specify how to learn the parameters
```

Then we fit it to data

```{r}

knn_fit_3 <- knn_spec_3 |> ## apply the data
  fit(species ~ ., data = p_sub_std_train)

```

Then we predict the test set

```{r}

p_sub_std_test <- p_sub_std_test |>
  mutate(knn3 = predict(knn_fit_3$fit,
                             p_sub_std_test, type = "prob")[,1])


p_sub_std_test$knn3
```


:::

#### 5. We will now analyse the fit of our three kNN models. Answer the questions below for this data.


a. For each model compute the confusion table, using whether the predictive probability was greater than 0.5 to label the observation.

::: unilur-solution

Model 1

```{r}
library(tidyr)

p_sub_std_test <- p_sub_std_test |>
  mutate(pred1 = ifelse(knn1 > 0.5, "Adelie", "Chinstrap"))
p_sub_std_test |> dplyr::count(species, pred1) |>  
  group_by(species) |>
  mutate(cl_err = n[pred1==species]/sum(n)) |>
  pivot_wider(names_from = pred1, 
              values_from = n,
              values_fill = 0) |>
  dplyr::select(species, Adelie, Chinstrap, cl_err)

```

Model 2

```{r}
library(tidyr)

p_sub_std_test <- p_sub_std_test |>
  mutate(pred2 = ifelse(knn2 > 0.5, "Adelie", "Chinstrap"))
p_sub_std_test |> dplyr::count(species, pred2) |>  
  group_by(species) |>
  mutate(cl_err = n[pred2==species]/sum(n)) |>
  pivot_wider(names_from = pred2, 
              values_from = n,
              values_fill = 0) |>
  dplyr::select(species, Adelie, Chinstrap, cl_err)

```

Model 3

```{r}
library(tidyr)

p_sub_std_test <- p_sub_std_test |>
  mutate(pred3 = ifelse(knn3 > 0.5, "Adelie", "Chinstrap"))
p_sub_std_test |> dplyr::count(species, pred3) |>  
  group_by(species) |>
  mutate(cl_err = n[pred3==species]/sum(n)) |>
  pivot_wider(names_from = pred3, 
              values_from = n,
              values_fill = 0) |>
  dplyr::select(species, Adelie, Chinstrap, cl_err)

```
:::

b. which value for $K$ seems best so far?

::: unilur-solution

Either $K = 5$ or $K = 20$ get better tets set accuracy than $K = 10$

:::

c. Compute the roc-auc for each value of $K$

::: unilur-solution

```{r}
#| echo: false

p_sub_std_test |>
  roc_auc(species, knn1) |>
  pull(.estimate)

p_sub_std_test |>
  roc_auc(species, knn2) |>
  pull(.estimate)

p_sub_std_test |>
  roc_auc(species, knn3) |>
  pull(.estimate)

```

:::


d. Plot the roc-auc curves for each value of $K$

::: unilur-solution

```{r}
#| echo: false

d_roc1 <- p_sub_std_test |>
  roc_curve(species, knn1) |>
  mutate(model = "K = 5") 

d_roc2 <- p_sub_std_test |>
  roc_curve(species, knn2) |>
  mutate(model = "K = 10") 

d_roc3 <- p_sub_std_test |>
  roc_curve(species, knn3) |>
  mutate(model = "K = 20") 



d_roc <- bind_rows(d_roc1, d_roc2, d_roc3)
ggplot(d_roc, aes(x=1-specificity, 
                  y=sensitivity, 
                  colour=model)) +
geom_path()
```

:::

e. Which value for $K$ seems better now


::: unilur-solution

While accuracy only looks at one cut-off, the roc-auc looks at all cutoff and provides a better idea of the accuracy of the predicted probabilities. 

Based on the roc-auc curves K = 20 appears to be the best model


:::

#### 6. Discuss with your neighbour, what you found the most difficult part of last week's content. Find some material (from resources or googling) together that gives alternative explanations that make it clearer. 

## `r emo::ji("wave")` Finishing up

Make sure you say thanks and good-bye to your tutor. This is a time to also report what you enjoyed and what you found difficult.
