---
title: "Week 3: Re-sampling and regularisation"
---

```{r}
week <- 3
source(here::here("course_info.R"))
```

## Main reference

```{r}
#| output: asis
schedule |>
  filter(Week == week) |>
  mutate(ref = glue::glue("[{Reference}]({Reference_URL})")) |>
  pull(ref)
```

## What you will learn this week

* Common re-sampling methods: bootstrap, cross-validation, permutation
* Cross-validation for parameter tuning
* Bootstrapping for understanding variance of parameter estimates
* Permutation to understand significance of associations between variables, and variable importance
* What can go wrong in high-d, and how to adjust using regularisation methods

## Lecture slides

- [html](https://iml.numbat.space/week3/slides.html)
- [pdf](https://iml.numbat.space/week3/slides.pdf)
- [qmd](https://iml.numbat.space/week3/slides.qmd)
- [R](https://iml.numbat.space/week3/slides.R)

## Tutorial instructions

Instructions:

- [html](https://iml.numbat.space/week3/tutorial.html)
- [qmd](https://iml.numbat.space/week3/tutorial.qmd)

## Assignments

```{r}
#| output: asis
show_assignments(week)
```
