---
title: "ETC3250/5250 Tutorial 4"
subtitle: "Re-sampling and regularisation"
author: "Prof. Di Cook"
date: "2024-03-11"
quarto-required: ">=1.3.0"
format:
    unilur-html:
        output-file: tutorial.html
        embed-resources: true
        css: "../assets/tutorial.css"
    unilur-html+solution:
        output-file: tutorialsol.html
        embed-resources: true
        css: "../assets/tutorial.css"
unilur-solution: true
---

```{r echo=FALSE}
# Set up chunk for all slides
knitr::opts_chunk$set(
  fig.width = 4,
  fig.height = 4,
  fig.align = "center",
  out.width = "60%",
  code.line.numbers = FALSE,
  fig.retina = 3,
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  dev.args = list(pointsize = 11)
)
```

```{r}
#| echo: true
#| code-fold: true
#| code-summary: "Load the libraries and avoid conflicts"
# Load libraries used everywhere
library(tidyverse)
library(tidymodels)
library(conflicted)
library(colorspace)
library(patchwork)
library(MASS)
library(randomForest)
library(gridExtra)
library(GGally)
library(geozoo)
library(mulgar)
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::select)
conflicts_prefer(dplyr::slice)
conflicts_prefer(palmerpenguins::penguins)
conflicts_prefer(tourr::flea)
```

```{r}
#| echo: false
# Set plot theme
theme_set(theme_bw(base_size = 14) +
   theme(
     aspect.ratio = 1,
     plot.background = element_rect(fill = 'transparent', colour = NA),
     plot.title.position = "plot",
     plot.title = element_text(size = 24),
     panel.background = element_rect(fill = 'transparent', colour = NA),
     legend.background = element_rect(fill = 'transparent', colour = NA),
     legend.key = element_rect(fill = 'transparent', colour = NA)
   )
)
```

## `r emo::ji("target")` Objectives

The goal for this week is for you to practice resampling methods, in order to tune models, assess model variance, and determine importance of variables. 

## `r emo::ji("wrench")` Preparation 

- Complete the quiz
- Do the reading related to week 3


## Exercises: 

Open your project for this unit called `iml.Rproj`. 

#### 1. Significance of PC coefficients

In the lecture, we used bootstrap to examine the significance of the coefficients for the second principal component from the womens' track PCA. Do this computation for PC1. Your question this time is: **Can we consider all of the coefficients to be equal?**

::: unilur-solution

 
:::


#### 6. Write a simple question about the week's material and test your neighbour, or your tutor. 

## `r emo::ji("wave")` Finishing up

Make sure you say thanks and good-bye to your tutor. This is a time to also report what you enjoyed and what you found difficult.
